Reviving the Lost Art of Combinator Graph Reduction

### 2022-12-06 experiments/superg/doc/paper/paper.tex
More on introduction.

\section{Introduction}
% Why did I research this. Give the motive, usually a knowledge gap or problem
Lazy evaluation of programs is a powerful idea.
It allows a program to operate on infinitely large data structures, and if part of the computation turns out to be unnecessary, it can be much more efficient than a strict evaluation.
In practice however, delaying computation often introduces overhead, making languages based on lazy evaluation slower than their strict counterparts.
To compete with a strict implementation, a lazy runtime must maximize lazyness while minizing overhead.
It is a difficult balance, and over the years many different \textit{abstract machines} have been proposed\cite{kieburtz_g-machine_1985,fairbairn_tim_1987,burn_spineless_1988,koopman_fresh_1989,jones_implementing_nodate}.
One of the early approaches to implementating lazy functional languages is combinator graph reduction\cite{turner_new_1979}.

@@

Combinator reduction was the runtime for David Turner's Miranda language\cite{turner_miranda_1985}, the popular choice for pure functional langauges at the time, and one of the few to be commercially supported.
Eventually the interest for Miranda and combinator graph reduction with it waned, and today Haskell is the dominant language.
Part of the reason may be that Miranda was closed source for 35 years, its source code was released only in 2020\cite{noauthor_open_2021}, but combinator reduction also deserves part of the blame.
Expressions are reduced in many small steps, so the per-step overhead greatly impacts runtime performance.
Every reduction step modifies the stack and usually also the graph, resulting in high memory traffic.
Another important issue is that the translation to combinators may result in much larger program.
For a lambda calculus expression of size $N$, the traditional bracket abstraction algorithm produces an output of $O(n^2)$ in the worst case, were other approaches such as supercombinators are linear.

These disadvantages have led to combinator graph reduction to be all but abandoned.
Recently however, a new algorithm for translation of lambda calculus expressions into combinatory logic has been proposed that is linear in the size of the input\cite{kiselyov_lambda_2018}.
While a clear improvement over previous translation strategies, it remains to be seen whether it can make the combinator reduction approach competitive with contemporary reduction engines.
This leads us to our first research question:

\textbf{RQ1:} How does Kiselyov's semantic translation compare to the classical bracket abstraction algorithm in terms of runtime performance of compiled programs?

We can take a set of benchmark programs and compile them with both algorithms, but to compare real-world execution time we also need a fast runtime to execute the compiled programs.

Alas, there is no performant combinator graph reduction engine on which we can execute programs compiled with Kiselyov's.
The now open sourced Miranda is a good starting point, but it is showing its age.
The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector.
Miranda's architecture is based on cells with a tag, head and tail as described in\cite{turner_new_1979}, but later work has shown that a tagless representation with direct pointers to code leads to better performance\cite{koopman_fresh_1989}.
In absence of an up to date reference on building a practical combinator reduction engine, we formlate the second research question:

@@

@@

\section*{Engine implementation}
\subsection*{Miranda}
\subsection*{TIGRE}

@@

\bibliographystyle{IEEEtran}
\bibliography{zotero}

>>>>

\section{Introduction}
% Why did I research this. Give the motive, usually a knowledge gap or problem
Lazy evaluation of programs is a powerful idea.
It allows a program to operate on infinitely large data structures, and if part of the computation turns out to be unnecessary, it can be much more efficient than a strict evaluation.
In practice however, delaying computation often introduces overhead, making languages based on lazy evaluation slower than their strict counterparts.
To compete with a strict implementation, a lazy runtime must maximize lazyness while minizing overhead.
It is a difficult balance, and over the years many different evaluator designs, or \textit{abstract machines}, have been proposed\cite{kieburtz_g-machine_1985,fairbairn_tim_1987,burn_spineless_1988,koopman_fresh_1989,jones_implementing_nodate}.
One of the early approaches to implementating lazy functional languages is combinator graph reduction\cite{turner_new_1979}.

@@

Combinator reduction was the runtime for David Turner's Miranda language\cite{turner_miranda_1985}, the popular choice for pure functional langauges at the time, and one of the few to be commercially supported.
Eventually the interest for Miranda and combinator graph reduction with it waned.
Part of the reason may be that Miranda was until very recently a closed source, licensed software (its source code was released only in 2020\cite{noauthor_open_2021}), leading to the creation of the open source Haskell language.
But combinator reduction also deserves part of the blame.
Firstly, expressions are reduced in many small steps, so the per-step overhead greatly impacts runtime performance.
Every reduction step modifies the stack and usually also the graph, resulting in high memory traffic.
Another important issue is that the translation to combinators may result in much larger programs.
For a lambda calculus expression of size $N$, the traditional bracket abstraction algorithm produces an output of $O(N^2)$ in the worst case, where other approaches such as supercombinators are linear\cite{spj_impl}.

These disadvantages have led to combinator graph reduction to be all but abandoned.
Recently however, a new algorithm for translation of lambda calculus expressions into combinatory logic has been proposed that is linear in the size of the input\cite{kiselyov_lambda_2018}.
While a clear improvement over previous translation strategies, it remains to be seen how big the impact is on real-world performance.
Is it enough to bridge the performance gap with newer reduction engines?
This leads us to our first research question:

\textbf{RQ1:} Does Kiselyov's semantic translation make combinator graph reduction competitive with contemporary lazy functional evaluators?

We use GHC as the reference for contemporary evaluator, and then measure the execution time of a set of benchmark programs.
We also need a fast reduction engine that can run the program compiled with Kiselyov's algorithm.
Alas, there is no performant combinator graph reduction engine tailored to executing such programs.
The now open sourced Miranda is a good starting point, but it is showing its age.
The 2020 port compiles cleanly on a recent GCC, but enabling optimizations is known to break the garbage collector.
Miranda's architecture is based on cells with a tag, head and tail as described in\cite{turner_new_1979}, but later work has shown that a tagless representation with direct pointers to code leads to better performance\cite{koopman_fresh_1989}.
In absence of an up to date reference on building a practical combinator reduction engine, we formlate the second research question:

@@

Concretely, our contributions are the following:
\begin{itemize}
    \item Kiselyov describes not one but two algorithms. The first does not have linear complexity, but tends to produce very compact code.
          The second is guaranteed to be of linear complexity, at the cost of being less compact in certain cases.
          We describe both algorithms in a way that is straightforward to implement.
    \item We implement a fast combinator graph reduction engine based on the ideas in Miranda~\cite{turner_miranda_1985} and later improvements~\cite{koopman_architecture_1992}.
    \item We evaluate the performance of Miranda, Haskell and our engine on a set of benchmark programs.
\end{itemize}

@@

\section*{Engine implementation}
\subsection*{Miranda}
Describe implementation of SASL/Miranda.

Compare performance with Open Source Miranda.

\subsection*{TIGRE}
Describe changes made to align with TIGRE.

Compare performance with SASL-style engine.

\section*{Performance evaluation}
Programs:
\begin{itemize}
    \item ackermann
    \item digits of e
    \item linfib
    \item lsort
    \item nfib
    \item primes
    \item queens
    \item tak
    \item towers of hanoi
    \item treesort
\end{itemize}

Evaluation performance of:

\begin{itemize}
    \item Bracket abstraction
    \item Kiselyov
    \item GHC
\end{itemize}

@@

\bibliographystyle{IEEEtran}
\bibliography{zotero, references}
